Type theories as quotient inductive-recursive types

### 2025-02-19 tex/main.tex
Some polishing Section 3

The equality constructor $[]\El$ is now accordingly replaced by the reduction
\begin{align}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\subTm{u}{\sigma}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution. The proof is similar to the proof of \cref{prop:local-confluence-1}, and can be found in \cref{appendix:proofs}.
%\begin{proposition}[Termination]
%  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
%  \danger
%\end{proposition}
%\begin{proof}
%  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
%\end{proof}
\begin{restatable}{proposition}{propConfluenceTwo}\label{prop:local-confluence-2}
  Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
  \qed
  \danger
\end{restatable}

@@

\begin{restatable}{proposition}{propCorrectness}\label{prop:correctness-1}
  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$. \qed
\end{restatable}

Finally, we have to show that both $\subTy{\blank}{\blank}$ and $\subTm{\blank}{\blank}$ are coherent.
%
\begin{restatable}{proposition}{propCoherence}\label{prop:coherence-2}
  For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$, the identities
  \danger
  \[
    \subTy{A}{\sigma} = \subTy{A}{\tau}
    \quad\text{and}\quad
    \subTm{t}{\sigma} = \subTm{t}{\tau}
  \]
  hold for any type $A$ and any term $t$. \qed
\end{restatable}

@@

\begin{restatable}{proposition}{propConfluenceThree}\label{prop:local-confluence-3}
  Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
  \qed
    \danger
\end{restatable}

Similar to term substitution, we can show that the lifted substitution $\sigma \uparrow A$ is propositionally equal to $\sigma^+$, based on two basic properties\footnote{%
  The reader is invited to prove them using the QII definition with transports.}
about the lifting.
\begin{lemma} \label{lem:lifting}
  The following statements hold:
  \[
    \idS^+ =^{\Sub\;(\Gamma, A)\;(\Gamma, A)} \idS
    \qquad\text{and}\qquad
    (\sigma; \tau)^+     =^{\Sub\;(\Gamma, \subTy{A}{\sigma; \tau})\;(\Theta, A)} \sigma^+ ; \tau^+.
  \]
\end{lemma}

\begin{restatable}{proposition}{propCorrectnessTwo}\label{prop:correctness-2}
    For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, the lifting
    $\sigma \uparrow A$ is propositionally equal to $\sigma^+$.
    \qed
\end{restatable}

As we are working with a different type theory from \cref{subsec:SC+U}, we still have to prove the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$ but the same argument still applies:
\begin{proposition} \label{prop:correctness-3}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$,
  the term $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$.
\end{proposition}

Finally, we show that these simultaneously defined functions are coherent.
\begin{restatable}{proposition}{propCoherenceUparrow} %[Coherence]
    For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$, we have
    \danger
    \[
      \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, \subTy{A}{\gamma})\;(\Delta, A)}_{\alert{\cc}} \tau \uparrow A
      \qquad\text{and}\qquad
      \subTm{t}{\sigma} =^{\lam{\gamma}\Tm\;\Gamma\;(\subTy{A}{\gamma})}_{\alert{\cc}} \subTm{t}{\tau},
    \]
    for any type $A$ and any term $t$.
    \qed
\end{restatable}

\subsubsection{Multiple lifting and the coherence of type substitution}

We also need to prove the coherence for type substitution, i.e., that $\subTy{A}{\sigma} = \subTy{A}{\tau}$ for every equality constructor $\cc : \sigma = \tau$.
To do this, we need to generalise the desired identity to account for liftings, since a naive induction does not work: for example, on $\Pi$-types, we have
\begin{align*}
  \subTy{(\Pi\;A\;B)}{\sigma} & = \Pi\;(\subTy{A}{\sigma})\;(\subTy{B}{\sigma\uparrow A}) \\
  \subTy{(\Pi\;A\;B)}{\tau} & = \Pi\;(\subTy{A}{\tau})\;(\subTy{B}{\tau\uparrow A})
\end{align*}
but the induction hypothesis does not give us $\subTy{B}{\sigma\uparrow A} = \subTy{B}{\tau\uparrow A}$.
Instead, we will define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ by a telescope $\Xi = A_1, \dots, A_n$, which intuitively means multiple applications of lifting: $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
Then, for each constructor $\cc : \sigma = \tau$, we show the identity
$\subTy{A}{\sigma \mathop{\upuparrows} \Xi} = \subTy{A}{\tau \mathop{\upuparrows} \Xi}$
for arbitrary $\Xi$, subsuming the special case $\subTy{A}{\sigma} = \subTy{A}{\tau}$ when $\Xi$ is empty.

@@

\begin{restatable}%[{Coherence for $[\blank]_{\Tel}$ and $\blank\upuparrows\blank$}]
{proposition}{coherenceTel}
  For every equality constructor $\cc : \sigma = \tau$ and telescope $\Xi$, we have
  $\subTel{\Xi}{\sigma} = \subTel{\Xi}{\tau}$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus \subTel{\Xi}{\sigma})\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$. \qed
  \danger
\end{restatable}

@@

\begin{corollary}
  We have $\subTy{A}{\sigma} = \subTy{A}{\tau}$ for every equality constructor $\cc : \sigma = \tau$.
  \qed
  \danger
\end{corollary}

>>>>

The equality constructor $[]\El$ is now accordingly replaced by the reduction
\begin{align}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\subTm{u}{\sigma}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution. The proof is similar to the proof of \cref{prop:local-confluence-1}, and can be found in \cref{appendix:proofs}.
%\begin{proposition}[Termination]
%  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
%  \danger
%\end{proposition}
%\begin{proof}
%  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
%\end{proof}
\begin{restatable}{proposition}{propConfluenceTwo}\label{prop:local-confluence-2}
  Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
  \danger
\end{restatable}

@@

\begin{restatable}{proposition}{propCorrectness}\label{prop:correctness-1}
  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $\sub{t}{\sigma}$ is propositionally equal to $\subTm{t}{\sigma}$.
\end{restatable}

Finally, we have to show that both $\subTy{\blank}{\blank}$ and $\subTm{\blank}{\blank}$ are coherent.
%
\begin{restatable}{proposition}{propCoherence}\label{prop:coherence-2}
  For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$, the identities
  \danger
  \[
    \subTy{A}{\sigma} = \subTy{A}{\tau}
    \quad\text{and}\quad
    \subTm{t}{\sigma} = \subTm{t}{\tau}
  \]
  hold for any type $A$ and any term $t$.
\end{restatable}

@@

\begin{restatable}{proposition}{propConfluenceThree}\label{prop:local-confluence-3}
  Type substitution $\subTy{\blank}{\blank}$ is terminating and locally confluent.
    \danger
\end{restatable}

The lifting $\sigma^+$ interacts well with identities and composition.
For comparision, the reader is invited to prove the corresponding QIIT version with transports of the following lemma.
%
\begin{lemma} \label{lem:lifting}
  For all contexts $\Gamma$, $\Delta$, $\Theta$, types $A : \Ty\,\Theta$ and substitutions $\sigma : \Sub\,\Gamma\,\Delta$ and $\tau : \Sub\,\Delta\,\Theta$, we have $\idS^+ =^{\Sub\;(\Theta, A)\;(\Theta, A)} \idS$ and $(\sigma; \tau)^+     =^{\Sub\;(\Gamma, \subTy{A}{\sigma; \tau})\;(\Theta, A)} \sigma^+ ; \tau^+$.
\end{lemma}

Using this lemma, we can show that the lifted substitution $\sigma \uparrow A$ agrees propositionally with its `default' reduction $\sigma^+$.

\begin{restatable}{proposition}{propCorrectnessTwo}\label{prop:correctness-2}
    For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, we have
    $\sigma \uparrow A = \sigma^+$.
\end{restatable}

As we are working with a different type theory from \cref{subsec:SC+U}, we still have to prove the identity $\sub{t}{\sigma} = \subTm{t}{\sigma}$ but the same argument still applies:
\begin{proposition} \label{prop:correctness-3}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Ty\;\Delta\;A$, we have $\sub{t}{\sigma} = \subTm{t}{\sigma}$.
\end{proposition}

Finally, we show that these simultaneously defined functions are coherent.
\begin{restatable}{proposition}{propCoherenceUparrow} %[Coherence]
    For every equality constructor $\cc : \sigma =^{\Sub\;\Gamma\;\Delta} \tau$, we have
    \danger
    \[
      \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, \subTy{A}{\gamma})\;(\Delta, A)}_{\alert{\cc}} \tau \uparrow A
      \qquad\text{and}\qquad
      \subTm{t}{\sigma} =^{\lam{\gamma}\Tm\;\Gamma\;(\subTy{A}{\gamma})}_{\alert{\cc}} \subTm{t}{\tau},
    \]
    for any type $A$ and any term $t$.
\end{restatable}

\subsubsection*{Multiple lifting and the coherence of type substitution}

We also need to prove the coherence for type substitution, i.e., that $\subTy{A}{\sigma} = \subTy{A}{\tau}$ for every equality constructor $\cc : \sigma = \tau$.
To do this, we need to generalise the desired identity to account for liftings, since a naive induction does not work: for example, on $\Pi$-types, we have
\[
  \subTy{(\Pi\;A\;B)}{\sigma} = \Pi\;(\subTy{A}{\sigma})\;(\subTy{B}{\sigma\uparrow A}) \qquad
  \subTy{(\Pi\;A\;B)}{\tau} = \Pi\;(\subTy{A}{\tau})\;(\subTy{B}{\tau\uparrow A})
\]
but the induction hypothesis does not give us $\subTy{B}{\sigma\uparrow A} = \subTy{B}{\tau\uparrow A}$.
Instead, we will define the \emph{multiple lifting} $\sigma \upuparrows \Xi$ by a telescope $\Xi = A_1, \dots, A_n$, which intuitively means multiple applications of lifting: $\sigma \uparrow A_1 \uparrow \dots \uparrow A_n$.
Then, for each constructor $\cc : \sigma = \tau$, we show the identity
$\subTy{A}{\sigma \mathop{\upuparrows} \Xi} = \subTy{A}{\tau \mathop{\upuparrows} \Xi}$
for arbitrary $\Xi$, subsuming the special case $\subTy{A}{\sigma} = \subTy{A}{\tau}$ when $\Xi$ is empty.

@@

\begin{restatable}%[{Coherence for $[\blank]_{\Tel}$ and $\blank\upuparrows\blank$}]
{proposition}{coherenceTel}
  For every equality constructor $\cc : \sigma = \tau$ and telescope $\Xi$, we have
  $\subTel{\Xi}{\sigma} = \subTel{\Xi}{\tau}$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus \subTel{\Xi}{\sigma})\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$.
  \danger
\end{restatable}

@@

\begin{corollary}
  We have $\subTy{A}{\sigma} = \subTy{A}{\tau}$ for every equality constructor $\cc : \sigma = \tau$.
  \danger
\end{corollary}
